Elephant Seal Age JAE Intraspecific Variation

### 2023-12-05 paper/paperv3.qmd
remove non-breeders from phenology plots

sealdat <- read_csv(here::here("data/raw/128L pull 2023_11_30.csv"), 
                    show_col_types = FALSE) %>% 
  mutate(observed = if_else(observed == "B", "Breeder", "Non-breeder"), 
         observed_int = if_else(observed == "Breeder", 1, 0), 
         pup_survived = pupseeneveragain > 0) %>% 
  filter(age > 3, 
         year > 1987)
```

@@

```{r}
pup_sex_mod <- glmer(as.factor(pupsex) ~ age + (1 | animalID) + (1 | year),
                     data = filter(sealdat, 
                                   pupsex %in% c("M", "F"), 
                                   age < 19), 
                     family = "binomial")
summary(pup_sex_mod)

@@

seal_phenology <- sealdat %>% 
  mutate(BA = firstobsbreeddoy, 
         BD = firstobsbreeddoy + breeddur, 
         MA = lastobsmoltdoy - moltdur, 
         MD = lastobsmoltdoy) %>% 
  drop_na(age, BA, BD, MA, MD)

@@

timing_summary <- seal_phenology %>% 
  group_by(age, observed) %>% 
  summarize(across(BA:MD, mean), 
            .groups = "drop") %>% 
  pivot_longer(cols = -c(age, observed),  
               names_to = "event", 
               values_to = "doy") %>% 
  group_by(age, observed) %>% 
  group_modify(\(data, keys) {
    tibble(event = c("Long trip", "Breeding", "Short trip", "Molting", "longtrip2"),
           doy = c(data$doy, 365) - c(0, data$doy))
  }) %>% 
  ungroup() %>% 
  mutate(event = factor(event, levels = c("Long trip", 
                                          "Breeding", 
                                          "Short trip", 
                                          "Molting", 
                                          "longtrip2")))
timing_summary_flipped <- timing_summary %>% 
  mutate(observed = ifelse(observed == "Breeder",
                           "Non-breeder",
                           "Breeder")) %>% 
  filter(event != "longtrip2")

@@

#Delete the non-breeders and add the dashed line as an average instead. 
#put panels next to each other 
#what do you do when breeding dates are in december? ifelse statement (if > 250), 200 - 365 (so use negative numbers if they are in december)
#show xy plots of each phenological variable against age in the supplemental and summary stats for the model in the supplement too 
#draw attention to the widths - 1) 2) 3) 
#assess linearity from plots, but double check visually and see if quadratic might fit better (we werent sure if age patterns would be linear or not, so we tested quadratic analyses throughout)

phen_plot1 <- ggplot(timing_summary, aes(x = age, y = doy, fill = event)) + 
  geom_area(position = position_stack(reverse = TRUE),
            color = NA) +
  geom_line(position = position_stack(reverse = TRUE),
            color = "white") +
  geom_line(data = timing_summary_flipped,
            position = position_stack(reverse = TRUE),
            color = "white",
            linetype = "dashed",
            alpha = 0.5) +
  scale_fill_discrete(type = c("#829BD1", "#E9B668", "#7EC6CC", "#E091B9", "#829BD1"),
                      breaks = c("Long trip", "Breeding", "Short trip", "Molting")) + 
  scale_x_continuous("Maternal Age (Years)", expand = expansion(mult = 0)) +
  scale_y_continuous("Day of Year", expand = expansion(mult = 0)) +
  facet_grid(cols = vars(observed)) +
  theme_few() + 
  theme(axis.title = element_text(size = 12),
        panel.spacing = unit(1, "lines"))

@@

# This data frame contains a model for each phenological variable w.r.t. age
phenology_models <- seal_phenology %>% 
  pivot_longer(BA:MD, names_to = "phenology", values_to = "doy") %>% 
  filter(observed_int > 0) %>% 
  select(phenology, doy, age, animalID, year) %>% 
  nest(model_data = c(doy, age, animalID, year)) %>% 
  mutate(model = map(model_data, fit_model),
         age_coef = extract_estimate(model, "age"),
         age_stderr = extract_stderr(model, "age"),
         age_pval = extract_pval(model, "age"))

#  mutate(observed = factor(observed, levels = c("Non-breeder", "Breeder"))) %>% 
         # age_signif = age_pval * 8 < 0.05,
         # breeder_coef = extract_estimate(model, "observedBreeder"),
         # breeder_stderr = extract_stderr(model, "observedBreeder"),
         # breeder_pval = extract_pval(model, "observedBreeder"),
         # breeder_signif = breeder_pval * 8 < 0.05,)

#remove bonferroni correction

@@

phen_plot2 <- phenology_models %>% 
  select(-c(model_data, model)) %>% 
  pivot_longer(c(ends_with("coef"), ends_with("stderr"), ends_with("signif")),
               names_to = c("param", "metric"),
               names_sep = "_",
               values_to = "value") %>% 
  pivot_wider(names_from = metric, values_from = value) %>% 
  mutate(signif = as.logical(signif), 
         param = factor(param, 
                        labels = c("'Delay with maternal age (days year'^-1 * ')'", 
                                   "'Delay with breeding (days)'")), 
         phenology = factor(phenology, 
                            labels = c("Breeding\narrrival", 
                                       "Breeding\ndeparture", 
                                       "Molt\narrrival", 
                                       "Molt\ndeparture"))) %>% 
  ggplot(aes(x = coef, y = phenology)) +
  # Prevent geom_rect() from establishing a continuous y-axis scale
  geom_point(alpha = 0) +
  geom_rect(aes(xmin = -Inf, xmax = Inf,
                ymin = event_begin, ymax = event_end,
                fill = event),
            phen_wrapped,
            inherit.aes = FALSE) +
  geom_pointrange(aes(xmin = coef - stderr, 
                      xmax = coef + stderr,
                      color = signif)) +
  scale_color_manual(values = c("FALSE" = "grey50", "TRUE" = "black"),
                     guide = "none") +
  scale_fill_discrete(type = c("#829BD1", "#E9B668", "#7EC6CC", "#E091B9", "#829BD1"),
                      breaks = c("Long trip", "Breeding", "Short trip", "Molting")) + 
  expand_limits(x = 0) +
  facet_grid(cols = vars(param), 
             scales = "free_x", 
             switch = "x",
             labeller = label_parsed) +
  theme_few() + 
  theme(axis.title = element_blank(),
        legend.direction = "horizontal",
        legend.title = element_blank(),
        strip.placement = "outside",
        strip.text = element_text(size = 12, color = "black"),
        panel.spacing = unit(1, "lines"))

plot_grid(
  get_legend(phen_plot2),
  phen_plot1 + theme(legend.position = "none"),
  phen_plot2 + theme(legend.position = "none"),
  align = "v",
  axis = "lr",
  labels = c("", "A", "B"),
  ncol = 1,
  rel_heights = c(0.1, 1, 1)
)

>>>>

sealdat <- read_csv(here::here("data/raw/128L pull 2023_11_30.csv"), 
                    show_col_types = FALSE) %>% 
  mutate(observed = if_else(observed == "B", "Breeder", "Non-breeder"), 
         observed_int = if_else(observed == "Breeder", 1, 0), 
         pup_survived = pupseeneveragain > 0,) %>% 
  filter(age > 3, 
         year > 1987)
```

@@

```{r}
#probably will need to take out animalID
pup_sex_mod <- glmer(as.factor(pupsex) ~ age + (1 | animalID) + (1 | year),
                     data = filter(sealdat, 
                                   pupsex %in% c("M", "F"), 
                                   age < 19), 
                     family = "binomial")
summary(pup_sex_mod)

@@

seal_phenology <- sealdat %>% 
  filter(observed == "Breeder") %>% 
  mutate(BA = firstobsbreeddoy, 
         BD = firstobsbreeddoy + breeddur, 
         MA = lastobsmoltdoy - moltdur, 
         MD = lastobsmoltdoy) %>% 
  drop_na(age, BA, BD, MA, MD)

mean_phenology <- seal_phenology %>% 
  summarize(across(BA:MD, mean), 
            .groups = "drop") %>% 
  pivot_longer(everything(),
               names_to = "event", 
               values_to = "doy")

@@

timing_summary <- seal_phenology %>% 
  group_by(age) %>% 
  summarize(across(BA:MD, mean), 
            .groups = "drop") %>% 
  pivot_longer(cols = -age,  
               names_to = "event", 
               values_to = "doy") %>% 
  group_by(age) %>% 
  group_modify(\(data, keys) {
    tibble(event = c("Long trip", "Breeding", "Short trip", "Molting", "longtrip2"),
           doy = c(data$doy, 365) - c(0, data$doy))
  }) %>% 
  ungroup() %>% 
  mutate(event = factor(event, levels = c("Long trip", 
                                          "Breeding", 
                                          "Short trip", 
                                          "Molting", 
                                          "longtrip2")))

@@

#Delete the non-breeders and add the dashed line as an average instead. 
#put panels next to each other 
#what do you do when breeding dates are in december? ifelse statement (if > 250), 200 - 365 (so use negative numbers if they are in december) **does not happen! no worries!!**
#show xy plots of each phenological variable against age in the supplemental and summary stats for the model in the supplement too 
#draw attention to the widths - 1) 2) 3) 
#assess linearity from plots, but double check visually and see if quadratic might fit better (we werent sure if age patterns would be linear or not, so we tested quadratic analyses throughout)

phen_plot1 <- ggplot(timing_summary, aes(x = age, y = doy, fill = event)) + 
  geom_area(position = position_stack(reverse = TRUE),
            color = NA) +
  geom_line(position = position_stack(reverse = TRUE),
            color = "white") +
  geom_hline(aes(yintercept = doy),
             mean_phenology,
             color = "white",
             linetype = "dashed",
             alpha = 0.5) +
  scale_fill_discrete(type = c("#829BD1", "#E9B668", "#7EC6CC", "#E091B9", "#829BD1"),
                      breaks = c("Long trip", "Breeding", "Short trip", "Molting")) + 
  scale_x_continuous("Maternal Age (Years)", expand = expansion(mult = 0)) +
  scale_y_continuous("Day of Year", expand = expansion(mult = 0)) +
  theme_few() + 
  theme(axis.title = element_text(size = 12),
        panel.spacing = unit(1, "lines"))

@@

# This data frame contains a model for each phenological variable w.r.t. age
phenology_models <- seal_phenology %>% 
  pivot_longer(BA:MD, names_to = "phenology", values_to = "doy") %>% 
  select(phenology, doy, age, animalID, year) %>% 
  nest(model_data = c(doy, age, animalID, year)) %>% 
  mutate(model = map(model_data, fit_model),
         age_coef = extract_estimate(model, "age"),
         age_stderr = extract_stderr(model, "age"),
         age_pval = extract_pval(model, "age"),
         age_signif = age_pval < 0.05)

@@

phen_plot2 <- phenology_models %>% 
  select(-c(model_data, model)) %>% 
  pivot_longer(c(ends_with("coef"), ends_with("stderr"), ends_with("signif")),
               names_to = c("param", "metric"),
               names_sep = "_",
               values_to = "value") %>% 
  pivot_wider(names_from = metric, values_from = value) %>% 
  mutate(signif = as.logical(signif), 
         # param = factor(param, 
         #                labels = "'Delay with maternal age (days year'^-1 * ')'"), 
         phenology = factor(phenology, 
                            labels = c("Breeding\narrrival", 
                                       "Breeding\ndeparture", 
                                       "Molt\narrrival", 
                                       "Molt\ndeparture"))) %>% 
  ggplot(aes(x = coef, y = phenology)) +
  # Prevent geom_rect() from establishing a continuous y-axis scale
  geom_point(alpha = 0) +
  geom_rect(aes(xmin = -Inf, xmax = Inf,
                ymin = event_begin, ymax = event_end,
                fill = event),
            phen_wrapped,
            inherit.aes = FALSE) +
  geom_pointrange(aes(xmin = coef - stderr, 
                      xmax = coef + stderr,
                      color = signif)) +
  scale_color_manual(values = c("FALSE" = "grey50", "TRUE" = "black"),
                     guide = "none") +
  scale_fill_discrete(type = c("#829BD1", "#E9B668", "#7EC6CC", "#E091B9", "#829BD1"),
                      breaks = c("Long trip", "Breeding", "Short trip", "Molting")) + 
  labs(x = expression('Delay with maternal age (days year'^-1 * ')')) +
  expand_limits(x = 0) +
  theme_few() + 
  theme(axis.title.y = element_blank(),
        legend.direction = "horizontal",
        legend.title = element_blank())

plot_grid(
  get_legend(phen_plot2),
  plot_grid(
    phen_plot1 + theme(legend.position = "none"),
    phen_plot2 + theme(legend.position = "none"),
    align = "h",
    axis = "tb",
    labels = c("A", "B"),
    nrow = 1
  ),
  ncol = 1,
  rel_heights = c(0.1, 1)
)
